\documentclass[../main.tex]{subfiles}
\begin{document}

\chapternotnumbered{Introduction}
\label{chap:Introduction}

The goal of this thesis is to investigate the use of certain numerical methods in simulating atmospheric flows around topography, the so-called \emph{orographic flows}. The text is structured as follows:

\begin{compactenum}
	\item chapter \Cref{chap:FLOWS} provides an introduction to the physics of \emph{orographic flows};
	\item chapter \Cref{chap:SPH} summarizes the main numerical method used, that is the \emph{Smoothed Particles (SPH)};
	\item chapter \Cref{chap:VORONOI} introduces a different numerical method: \emph{Semi-Implicit Lagrangian Voronoi Approximation (SILVA)};
	\item chapter \Cref{chap:EXPERIMENTS} presents these numerical methods through numerical experiments and discusses the obtained results.
\end{compactenum}

\section*{Motivation}

Earth's atmosphere is one of the most thoroughly studied and well-described natural systems. With the development of computer architecture, a great effort has been made to complete the \emph{experimental} data with the \emph{predictive} abilities of computational modelling. The need for effective and reliable predictive power is naturally well-founded -- many processes in the atmosphere have a massive influence on the lives of Earth's inhabitants.

One example of such a process is orographic flow -- an atmospheric flow of air mass over topography. Such flows are fairly common in nature and most importantly they give rise to effects of high relevance in atmospheric dynamics \& weather forecasting, such as

\begin{compactenum}
	\item downslope windstorms;
	\item \textcolor{gray}{atmospheric} internal gravity waves\sidenotemark;
	\item clear air turbulence.
\end{compactenum}

\sidenotetext{Note the distinction between \emph{gravitational waves} -- disturbances of \emph{space-time} studied in the general theory of relativity, and \emph{gravity waves} -- disturbances of \emph{a fluid} studied mainly in meteorology and oceanology. To make this distinction even more evident, the name \emph{internal} gravity waves is sometimes used in the literature.}

The ability to obtain accurate and reliable \emph{numerical} solutions to such flows is thus highly \emph{practically relevant. \sidenotemark} \sidenotetext{And scientifically relevant also: the study of orographic flows and mainly internal gravity waves is an active field in the atmospheric science community.} In this work, we aim to explore the suitability of SPH and SILVA for this task.


\end{document}
